Accelerating Optimal Power Flow: Condensed-Space Interior-Point Methods and Automatic Differentiation on GPUs

### 2023-07-20 paper/main.tex
code directory added

\input{preamble}
\title{Accelerating Optimal Power Flow: Condensed-Space Interior-Point Methods and Automatic Differentiation on GPUs}

@@

\date{\small
  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
}
\begin{document}
\maketitle

While graphics processing units (GPUs) have showcased impressive capabilities in various computing domains, their utilization in large-scale constrained nonlinear optimization regimes, such as alternating current (AC) optimal power flow (OPF) problems, has been somewhat limited. This limitation stems from the challenges associated with parallel factorization of indefinite sparse matrices commonly encountered within constrained optimization algorithms (Anitescu et. al. 2021). Although GPU computation can accelerate various other parts of the optimization process, including automatic differentiation (AD) and sparse matrix-vector multiplications, the slow data transfer between host and device memory hinders the ad-hoc implementation of GPU accelerations. To fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework on the GPU that incorporates AD, linear algebra, and optimization while minimizing data transfers to and from host memory.

@@

The paper will present comprehensive numerical benchmark results to demonstrate the efficiency of our proposed approach. Our method is implemented in our packages, a nonlinear optimization solver MadNLP.jl and an automatic differentiation tool SIMDiff.jl, with the solution of the KKT system being carried out using the external CUSOLVER library. We compare our method against the standard CPU approach (W\"{a}chter and Beigler, 2006) and the recently developed reduced-space interior-point method on GPUs (Pacaud et. al., 2022) using the standard data available in pglib-opf. Preliminary results indicate that our proposed framework has the potential to accelerate the solution of AC OPF problems up to moderate tolerances ($10^{-6}$) by an order of magnitude compared to existing tools such as Ipopt (interfaced with MATPOWER or PowerModels.jl), especially for large-scale instances. We will conclude with a discussion on the future extensibility of our method for more complex optimization tasks, such as multi-period, security-constrained, and joint transmission-distribution optimization.

% \bibliographystyle{plain}
% \bibliography{main}

>>>>

% \input{preamble}
\title{Accelerating Optimal Power Flow: Condensed-Space Interior-Point Methods and Automatic Differentiation on GPUs}

@@

\date{\small
  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
}
\begin{document}
\maketitle
\begin{abstract}
This paper presents a computational framework for solving AC optimal
power flow (OPF) problems using GPUs.  Despite the impressive
capabilities of GPUs in various computing domains, their application
in AC OPF has been limited by various challenges, including sparse
matrix factorization and automatic differentiation.  To address these
issues, we propose two strategies: (i) a condensed-space
interior-point method (IPM) with inequality relaxation, and (ii) a
single-instruction, multiple-data (SIMD) abstraction of nonlinear
programs.  By relaxing equality constraints and condensing the
Karush-Kuhn-Tucker system to the primal space, we enable efficient
sparse matrix factorization on GPUs without numerical pivoting.
Further, we introduce the SIMD abstraction for parallelized derivative
evaluations on GPUs.  The paper provides numerical benchmark results,
comparing our approaches with state-of-the-art tools such as JuMP,
AMPL, and Ipopt, demonstrating an order of magnitude 
speedup.  The paper concludes by discussing the future extensibility.
\end{abstract}

\section{Introduction}
While graphics processing units (GPUs) have showcased impressive
capabilities in various computing domains, their utilization in
large-scale constrained nonlinear optimization regimes, such as
alternating current (AC) optimal power flow (OPF) problems, has been
somewhat limited. This limitation stems from the challenges associated
with parallel factorization of indefinite sparse matrices commonly
encountered within constrained optimization algorithms
\cite{anitescu2021targeting}. Although GPU computation can accelerate
various other parts of the optimization process, including automatic
differentiation (AD) and sparse matrix-vector multiplications, the
slow data transfer between host and device memory hinders the ad-hoc
implementation of GPU accelerations. To fully leverage the
capabilities of modern GPU hardware, it is essential to implement a
comprehensive computational framework on the GPU that incorporates AD,
linear algebra, and optimization while minimizing data transfers to
and from host memory.

@@

We present comprehensive numerical benchmark results to demonstrate
the efficiency of our proposed approach. Our method is implemented in
our packages, a nonlinear optimization solver MadNLP.jl and an
automatic differentiation tool SIMDiff.jl, with the solution of the
KKT system being carried out using the external CUSOLVER library. We
compare our method against the standard CPU approach (W\"{a}chter and
Beigler, 2006) and the recently developed reduced-space interior-point
method on GPUs (Pacaud et. al., 2022) using the standard data
available in pglib-opf. Preliminary results indicate that our proposed
framework has the potential to accelerate the solution of AC OPF
problems up to moderate tolerances ($10^{-6}$) by an order of
magnitude compared to existing tools such as Ipopt (interfaced with
MATPOWER or PowerModels.jl), especially for large-scale instances. We
will conclude with a discussion on the future extensibility of our
method for more complex optimization tasks, such as multi-period,
security-constrained, and joint transmission-distribution
optimization.

\paragraph*{Contributions}
The contribution of this paper is three-fold. First, we present a condensed-space interior point method with inequality relaxation strategy is a general framework for solving sparse, large-scale nonlinar optimization problems on GPUs. Second, we present a SIMD abstraction of nonlinear program, which facilitates the implementation of parallel automatic differentiation on GPUs. Third, we present a comprehensive numerical benchmark result for the proposed two strategy using the standard test cases, thereby demonstrating the effectiveness of the two proposed methods. To the best of our knowledge, our approach is the fastest algorithm to solve AC OPF problems to $10^{-5}$ accuracy.

\paragraph*{Organization} The paper is organized as follows. In the remainder of the current section, we introduce the mathematical notation. In Section \ref{sec:prelim}, we provide general preliminary knowledge on numerical optimization and GPU computing. In Section \ref{sec:simd}, we present the SIMD abstraction strategy for large-scale nonlinear programs and their advantages in terms of implementing parallel automatic differentiation. Section \ref{sec:ipm} presents the optimization algorithm under study, the condensed-space interior point method wiht inequality relaxation strategy. Section \ref{sec:num} presents the numerical results, comparing our approach with other state-of-the-art solution methods on GPUs. Finally, conclusions and future outlooks are given in Section \ref{sec:conc}.

\paragraph*{Notation}
We denote the set of real numbers and the set of integers by $\mathbb{R}$ and $\mathbb{I}$. We let $[M]:=\{1,2,\cdots,M\}$.

\section{Preliminaries}\label{sec:prelim}
This section introduces two key background information: GPU computing and numerical optimization for large-scale nonlinear optimzation problems.

\subsection{GPU Computing}\label{sec:gpu}

GPUs are particularly effective in performing the following operations:
\begin{subequations}\label{eqn:pattern}
  \begin{align}
    y&\leftarrow \left[g(x; q_j)\right]_{j\in [J]}\\
    y&\leftarrow y + \sum_{k\in [K]} h(x;s_k)\\
    o&\leftarrow \sum_{i\in [I]} f(x; p_i),
  \end{align}
\end{subequations}

\subsection{Numerical Optimization}\label{sec:numopt}
The state-of-the-art solution strategies for AC OPF problems rely on numerical optimziation software, either general-purpose \cite{} or specialized \cite{}.
Three main pillars of modern nonlinear optimization software include not only optimization solver, but also automatic differentiation tool (often provided by or interfaced with algebraic modeling language) and linear solvers. In fact, most of the expensive computational efforts are delegated to the external linear solver and automatic differentiation library. Optimization solver orchestrates the operation of these tools to drive the solution iterate to the stationary point of the optimization problem.

\paragraph{Optimization Solvers}
Optimization solvers

\paragraph{Automatic Differentation}
For the efficient solution of the optimization problems, it is crucial to be have the ability to

\paragraph{Linear Solvers}

\begin{figure}
  \centering 
  \begin{tikzpicture}[remember picture, scale=.9, font=\small]
    \node[draw,align=left] (I) at (0,2.75) {
      $\bullet$ Instruction $g(\cdot,\cdot)$\\
      $\bullet$ Pointer to array $x$
    };
    \node[draw] at (-4,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (A\x) at (0,-\x/2) {$s_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \node[draw] at (0,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \node[draw] at (4,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (C\x) at (0,-\x/2) {$y_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \foreach\x in {1,2,3,4,5,7}{
      \draw[->] (A\x.east) -- (B\x.west);
      \draw[->] (B\x.east) -- (C\x.west);
      \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
    }
    \node at (-4,-2.5) {Input Data};
    \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
    \node at (4,-2.5) {Output Data};
  \end{tikzpicture}
  \label{fig:simd}
  \caption{A schematic description of SIMD parallelism}
\end{figure}

\section{SIMD Abstraction of Nonlinear Programs}\label{sec:simd}
\begin{subequations}\label{eqn:prob}
  \begin{align}
    \min_{x^L\leq x \leq x^U}& \sum_{\ell\in[L]}\sum_{i\in [I_k]} f^{(\ell)}(x; p^{(\ell)}_i)\\
    \st\; &\forall m\in[M]:\\
    &\begin{aligned}[t]
      g^{(m)}_\flat\leq &\left[g^{(m)}(x; q_j)\right]_{j\in [J]} \\
      &+\sum_{n\in [N_m]}\sum_{k\in [K_n]}h^{(n)}(x; s^{(n)}_{k}) \leq g^{(m)}_\sharp,
    \end{aligned}
  \end{align}
\end{subequations}
\section{Condensed-Space Interior Point Methods}\label{sec:ipm}

\section{Numerical Results}\label{sec:num}

\section{Conclusions and Future Outlook}\label{sec:conc}

\bibliographystyle{IEEEtran}
\bibliography{main}
